---
title: "class15inclass"
author: "Nicole Jacobson"
date: "11/17/2021"
title: "Class 11 in class"
output: pdf_document
author: Nicole Jacobson A13081206
date: 211117
---

```{r}
library(BiocManager)
library(DESeq2)

```

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

```{r}
head(counts)
```

```{r}
head(metadata)
```

Let's check the correspondence of the metadata and count data setup. 

```{r}
metadata$id
```
```{r}
colnames(counts)
```

```{r}
# we can use "==" to see if two arguments are equal
metadata$id == colnames(counts)
```

```{r}
all(metadata$id == colnames(counts))
```

# We want to start by comparing the contorl vs treated groups.
  We will need to access all the control columns in the data.
  
```{r}
control.inds <- metadata$dex == "control"
control.ids <- metadata[control.inds,]$id
```

Will use these IDs to access just the control columns of our "count" data

# Q2. How many ‘control’ cell lines do we have?
4

```{r}
head(counts[,control.ids])
```


# Q1. How many genes are in this dataset?
38694

```{r}
nrow(counts)
```

There are `r nrow(counts)` rows/genes in this dataset


```{r}
control.mean <- (rowMeans(counts[,control.ids]))
head(control.mean)
```
Do the same for drug treated

```{r}
treated.inds <- metadata$dex == "treated"
treated.ids <- metadata[treated.inds,]$id
```

```{r}
head(counts[,treated.ids])
```

```{r}
treated.mean <- (rowMeans(counts[,treated.ids]))
head(treated.mean)
```

# Compare the control and treated with a quick plot of contorl vs. treated mean counts

```{r}
plot(control.mean, treated.mean)

```

We need to log transform this data!! log log scale

```{r}
plot(log="xy", treated.mean, control.mean)
```

```{r}
log2(20/20)
```

```{r}
log2(80/20)
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

fc for fold change. This command will add another column called log2fc

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

```{r}
plot(meancounts$log2fc)
```

# Now we want to drop all the 0.00 counts

```{r}
head(meancounts[,1:2])
```

```{r}
head(meancounts[,1:2]==0)
```

The `which()` function tells us the TRUE results in a logical vector.

  But, it's not that helpful since it just gives us the coorinate of the column we're looking for. 
  
```{r}
head(which(meancounts[,1:2]==0))
```

Now we can try and exclude the zeros. Here we only care about the rows (if there is a zero in any column we want to exclude it)
```{r}
inds <- which(meancounts[,1:2] ==0, arr.ind=TRUE)
head(inds)
```

```{r}
# these are the rows I want to remove
to.rm <- unique((inds[,"row"]))
```


```{r}
mycounts <- (meancounts[-to.rm,])
```

Now we have `r nrow(mycounts)` genes left

```{r}
nrow(mycounts)
```

Now we are interested in how many genes are up/down regulated at the log2 fold change level.

```{r}
# this is how many genes are unregulated
sum(mycounts$log2fc > +2)

```
what percent is this of the total genes?

```{r}
# this will give us what percent are upregulated. 1.15% are upregulated
(sum(mycounts$log2fc > +2) / (nrow(mycounts)))*100
```

```{r}
# this will give us what percent of the genes are downregulated. 1.68%
(sum(mycounts$log2fc < -2) / (nrow(mycounts)))*100
```

```{r}
library(DESeq2)
citation("DESeq2")
```
# we first need to setup the DESeq input dataset.

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```
# run the DESeq analysis pipeline

```{r}
dds <- DESeq(dds)
```


```{r}
res <- results(dds)
head(res)
```


# let's start with a volcano plot. 

  This is a very common data visualization for this kind of data.


```{r}
plot(res$log2FoldChange, -log(res$padj))
abline(v=c(-2,2), col="green")
abline(h=-log(0.05), col="purple")

```


# Adding annotation data

  We want to add meaningful gene names to our dataset so we can make some sense fo what is going on.
  
  For this we will use two bioconductor packages, one does the work and is called annotationdbi and the other contains data we are going to map and     it's called org.hs.eg.db

```{r}
# BiocManager::install("AnnotationDbi")
```

```{r}
# BiocManager::install("org.Hs.eg.db")
```


```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```


```{r}
columns(org.Hs.eg.db)
```

Here we map to "SYMBOL" the common gene name that the world understands

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
head(res$symbol)

```

```{r}
head(res)
```







