---
title: "functionsintro"
author: "Nicole Jacobson A13081206"
Date: 211015
output: pdf_document
html_document: null
---

## R markdown intro

# **Intro to  functions**

```{r}
rescale <- function(x) {
  rng <- range(x) 
(x-rng[1])/(rng[2]-rng[1])
}

rescale
rescale(1:10)

rescale(c(1,4,10,20))
```


> Q1. Write a function grade() to determine an overall grade from a vector of student homework
assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
NA value) this can be used as a score to be potentially dropped. Your final function should be
adquately explained with code comments and be able to work on an example class gradebook
such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```



```{r}
# tells us the position at which the minimum value is found
which.min(student1)
```

```{r}
# - in the front will do everything but that is listed
# add mean at the front to get the mean of those values
mean(student1[ - which.min(student1)])


```


```{r}
# Gives us the position where we find an NA
which(is.na(student2)) 

```

```{r}
# Let's replace the NAs with zeros.using the [] we are making the positions with an NA 0
student.prime <- student3
student.prime[is.na(student.prime)] = 0
student.prime


```


```{r}
#put pieces together to get the final function
mean(student.prime[ - which.min(student.prime)])

```


```{r}
# now we can make our function. We will take out working piece and make it a function (code -> extract function)
grade <- function(student.prime) {
  student.prime[is.na(student.prime)] = 0
  mean(student.prime[ - which.min(student.prime)])
}
```


```{r}
grade(student1)
grade(student2)
grade(student3)
```

```{r}
#' Calculate the average score for a class of scores excluding the lowest score. Missing values are treated as zero scores
#'
#' @param student.prime numeric vector of homework scores
#'
#' @return average score
#' @export 
#'
#' @examples student <- c(100, NA, 90, 80)
#' grade(student)
grade <- function(student.prime) {
  # made the NA values equal to zero
  student.prime[is.na(student.prime)] = 0
  # made to exclude the lowest value
  mean(student.prime[ - which.min(student.prime)])
}



```



```{r}
# now we can take the grade book and grade the class (example class grade book found here https://tinyurl.com/gradeinput

url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names=1)
gradebook
```

```{r}
# apply(x, margin, fun, ...) x is the dataset, margin 1 indicates rows 2 indicates columns or both c(1,2), fun is the functiom
apply(gradebook, 1, grade)

```

```{r}
# who is the top scoring student?
results <- apply(gradebook, 1, grade)
sort (results, decreasing = TRUE)
```

```{r}
# another option
which.max(results)
```

```{r}
# was was the toughest assignment? Which column has the lowest score? Two options.
hw.mean <- apply(gradebook, 2, mean, na.rm = TRUE)
which.min(hw.mean)

hw.median <- apply(gradebook, 2, median, na.rm = TRUE)
which.min(hw.median)

```

```{r}
boxplot(gradebook)
```

```{r}
grade2 <- function(student.prime, drop.lowest=TRUE) 
  {student.prime <- as.numeric(student.prime)
student.prime[is.na(student.prime)] = 0 
if(drop.lowest) 
{mean(student.prime[ -which.min(student.prime)])
} else {
    mean(student.prime)}
}

```














